# MPlus Brownbag Multilevel Modeling Tutorial

## Data:

```{rdata, comment=NA,message=FALSE }
load("~/RStudioProjects/MplusTutorials/MplusMultilevel/data/pop2.Rdata")
library(psych)
library(xtable)
library(arm)
library(car)
library(texreg)
library(mice)
str(pop2,give.attr=FALSE)
```


```{r descriptives, results='asis'}

print(xtable(describe(pop2[,1:6],interp=F,skew=F)),type="html")

```

## Add Unique ID variable

```{r add ID, comment=NA}
pop2$id <-(pop2$class*100) + pop2$pupil
u.id <- unique(pop2$id)
print(paste("# unique id variables =",length(u.id)))
```


```{r descriptives2, results='asis'}

print(xtable(describe(pop2[,c(1:6,16)],interp=F,skew=F)),type="html")
```

```{r, head, comment=NA}
some(pop2[,c(2,1,16)])

```

## historgrams

```{r hists, }
par(mfcol=c(2,2))
hist(pop2$popular,col="red",breaks=10)
hist(pop2$sex,col="red",breaks=2)
hist(pop2$extrav,col="red",breaks=10)
hist(pop2$texp,col="red",breaks=10)
```

## correlations

```{r corrplot}
corrplot(pop2[,3:6],color=TRUE)
```


## Research Question:


Are extraverted students more popular than intraverted students?


One OLS regression, ignoring nesting within classes:
$$
Y_{i}=\beta_{0} + \beta_{1} X_{1i} + \beta_{2} X_{2i} + e_{i}
$$

$$
popularity_{i}=\beta_{0} + \beta_{1} sex_{1i} + \beta_{2} extraversion_{2i} + e_{i}
$$

separate OLS regression equation for each class:
$$
Y_{ij}=\beta_{0j} + \beta_{1j}X_{1ij} + \beta_{2j}X_{2ij} + e_{ij}
$$
$$
popularity_{ij}=\beta_{0j} + \beta_{1j}sex_{1ij} + \beta_{2j}extraversion_{2ij} + e_{ij}
$$

```{r olsmods, results='asis'}
cp.mod <- lm(Zpopular ~ Zsex + Zextrav ,pop2)
np.mod <- lm(Zpopular ~ Zsex + Zextrav + class,pop2)
ml.mod <- lmer(Zpopular ~ Zsex + Zextrav + (1|class),pop2)
mod.names <- c("1 OLS","100 OLS","MLM")
htmlreg(list(cp.mod,np.mod,ml.mod),omit.coef="class",model.names=mod.names)
```


Intercept only model:

level 1: 
$$
Y_{ij}=\beta_{0j}+e_{ij}
$$

level 2:
$$
\beta_{0j}=\gamma_{00}+u_{0j}
$$

Intraclass correlation (ICC):
$$ 
\rho =\frac{\sigma_{u_0}^2}{\sigma_{u_0}^2 + \sigma_{e}^2}
$$


```{r mlmods, results='asis', comment=NA, message=FALSE}
mod0 <- lmer(popular ~ 1 + (1 |class), pop2,REML=FALSE)
mod1 <- lmer(popular ~ sex + extrav + (1 |class), pop2,REML=FALSE)
mod2 <- lmer(popular ~ sex + extrav + texp + (1 + extrav|class),
             pop2,REML=FALSE)
mod3 <- lmer(popular ~ sex + extrav*texp + (1+ extrav|class),
             pop2,REML=FALSE)
mod.names <- c("unconditional","level 1","level 2","interaction")
htmlreg(list(mod0,mod1,mod2,mod3),model.names=mod.names,symbol="+",
        caption="Multilevel Models",caption.above=TRUE)
```


## Preparing pop2 data for Mplus in R
```{r mplusdata, eval=FALSE}
# The foreign package contains many functions to import
# various types of data from other stats programs.
library(foreign)
# use this command to read in an SPSS file:
pop2 <- read.spss(file="data/popular2.sav") 
# The following command creates a .dat file and
# and .inp file with the basic Mplus code.
library(MplusAutomation)
prepareMplusData(df=popular2,
                 filename="data/popular2.dat",      
                 inpfile="mplus/popular2.bare.inp")
```

