# MPlus Brownbag Multilevel Modeling Tutorial

## Data:

```{rdata, comment=NA,message=FALSE,warning=FALSE}
load("C:/RStudioProjects/MplusTutorials/MplusMultilevel/data/pop2.Rdata")
library(psych)
library(xtable)
library(arm)
library(car)
library(texreg)
library(GGally)
str(pop2,give.attr=FALSE)
```


```{r descriptives, results='asis'}

print(xtable(describe(pop2[,1:6],interp=F,skew=F)),type="html")

```

## Change 'sex' to 'female' and add Unique ID variable 

```{r add ID, comment=NA}
names(pop2)[4] <- "female"
pop2$id <-(pop2$class*100) + pop2$pupil
u.id <- unique(pop2$id)
print(paste("# unique id variables =",length(u.id)))
```


```{r descriptives2, results='asis'}

print(xtable(describe(pop2[,c(1:6,16)],interp=F,skew=F)),type="html")
```

```{r, head, comment=NA}
some(pop2[,c(2,1,16)],n=20)

```

## historgrams

```{r hists, }
par(mfcol=c(2,2))
hist(pop2$popular,col="red",breaks=10)
hist(pop2$female,col="red",breaks=2)
hist(pop2$extrav,col="red",breaks=10)
hist(pop2$texp,col="red",breaks=10)
```

## correlations

```{r corrplot}
# corrplot is found in the 'arm' package
corrplot(pop2[,3:6],color=TRUE)
```

```{r pairplot, cache=TRUE}
ggpairs(pop2[,3:6])
```


## Research Question:


### Are extraverted students more popular than intraverted students?
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />

One OLS regression, ignoring nesting within classes (n=2000):
$$
Y_{i}=\beta_{0} + \beta_{1} X_{1i} + \beta_{2} X_{2i} + e_{i}
$$
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
$$
popularity_{i}=\beta_{0} + \beta_{1} female_{1i} + \beta_{2} extraversion_{2i} + e_{i}
$$

<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
separate OLS regression equation for each class:
$$
Y_{ij}=\beta_{0j} + \beta_{1j}X_{1ij} + \beta_{2j}X_{2ij} + e_{ij}
$$
<br />
<br />
<br />
<br />
<br />

$$
popularity_{ij}=\beta_{0j} + \beta_{1j}female_{1ij} + \beta_{2j}extraversion_{2ij} + e_{ij}
$$
<br />
<br />
<br />
<br />
<br />

 -or-
 $$
Y_{ij}=\beta_{0j} + \beta_{1j}X_{1ij} + \beta_{2j}X_{2ij} + \beta_{3j}X_{3j} + ... + e_{ij}
$$
<br />
<br />
<br />
<br />
<br />
 $$
popularity_{ij}=\beta_{0j} + \beta_{1j}female_{1ij} + \beta_{2j}extraversion_{2ij}  + \beta_{3j}class_{3j} + ... + e_{ij}
$$
<br />
<br />
<br />
<br />
<br />

```{r olsmods, results='asis'}


# Models
ols <- lm(popular ~ female + extrav, pop2)
ols.classIV.mod <- lm(popular ~ female + extrav + factor(class),pop2)
ml.mod <- lmer(popular ~ female + extrav + (1|class),pop2)
mod.names <- c("OLS","OLS w/class as IV","MLM")

# make table
htmlreg(list(ols,ols.classIV.mod,ml.mod),omit.coef="class",model.names=mod.names)

# show class coefficients

ols.classIV.noInt.mod <- lm(popular ~ factor(class) + female + extrav - 1,pop2)
class.coefs <- as.data.frame(cbind(coef(ols.classIV.noInt.mod)[c(1:20)],
                                   se.coef(ols.classIV.noInt.mod)[c(1:20)]))
names(class.coefs) <- c("coefficient","s.e.")
print(xtable(class.coefs,caption="First 20 class coefficients"),type="html")
paste("Range of class coefficients = ",round(range(coef(ols.classIV.noInt.mod)[1:100])[1],2)," to ",round(range(coef(ols.classIV.noInt.mod)[1:100])[2],2) )
paste(" mean s.e. of class coefficients = ", round(mean(se.coef(ols.classIV.noInt.mod)[1:100]),2))
```


Intercept only model:

level 1: 
$$
Y_{ij}=\beta_{0j}+e_{ij}
$$
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
level 2:
$$
\beta_{0j}=\gamma_{00}+u_{0j}
$$
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />
<br />

combined:
$$
Y_{ij}=\gamma_{00}+u_{0j} + e_{ij}
$$
<br />
<br />
<br />
<br />
<br />
Intraclass correlation (ICC):
$$ 
\rho =\frac{\sigma_{u_0}^2}{\sigma_{u_0}^2 + \sigma_{e}^2}
$$


```{r mlmods, results='asis', comment=NA, message=FALSE}
mod0 <- lmer(popular ~ 1 + (1 |class), pop2,REML=FALSE)
mod1 <- lmer(popular ~ female + extrav + (1 |class), pop2,REML=FALSE)
mod2 <- lmer(popular ~ female + extrav + texp + (1 + extrav|class),
             pop2,REML=FALSE)
mod3 <- lmer(popular ~ female + extrav*texp + (1+ extrav|class),
             pop2,REML=FALSE)
mod4 <- lmer(popular ~ female + Cextrav*Ctexp + (1 + Cextrav |class),pop2, REML=F)
names(mod4@fixef)[3:5] <- c("extrav","texp","extrav:texp")
mod.names <- c("unconditional","level 1","level 2","interaction","int. vars. centered")
htmlreg(list(mod0,mod1,mod2,mod3,mod4),model.names=mod.names,symbol="+",
        caption="Multilevel Models",caption.above=TRUE)
```


## Preparing pop2 data for Mplus in R
```{r mplusdata, eval=FALSE}
# The foreign package contains many functions to import
# various types of data from other stats programs.
library(foreign)
# use this command to read in an SPSS file:
popular2 <- read.spss(file="data/popular2.sav",
                  use.value.labels=FALSE,
                  to.data.frame=TRUE
                  ,use.missings=TRUE) 
# The following command creates a .dat file and
# and .inp file with the basic Mplus code.
library(MplusAutomation)
prepareMplusData(df=popular2,
                 filename="data/pop2.dat",      
                 inpfile="mplus/bare.inp")
```

